Synthesizing Random Generators via Higher-Order Representations

### 2019-01-17 paper/sections/hrep.tex
Unfolded figure
WIP

What to say here:

\begin{itemize}
\item Haskell ADTs
\item Type driven generation
\item Limitations
\item Key idea: Higher order representations
\item Canonical representation of the data type
\item Pattern matchings representation
\item Abstract interface representation
\item Composing representations
\end{itemize}

\begin{figure}[b]
  \centering
  \input{tikz/hrep.tex}
  \caption{The idea}
\end{figure}

Pellentesque dapibus suscipit ligula. Donec posuere augue in quam. Etiam vel
tortor sodales tellus ultricies commodo. Suspendisse potenti. Aenean in sem ac
leo mollis blandit. Donec neque quam, dignissim in, mollis nec, sagittis eu,
wisi. Phasellus lacus. Etiam laoreet quam sed arcu. Phasellus at dui in ligula
mollis ultricies. Integer placerat tristique nisl. Praesent augue. Fusce
commodo. Vestibulum convallis, lorem a tempus semper, dui dui euismod elit,
vitae placerat urna tortor vitae lacus. Nullam libero mauris, consequat quis,
varius et, dictum id, arcu. Mauris mollis tincidunt felis. Aliquam feugiat
tellus ut neque. Nulla facilisis, risus a rhoncus fermentum, tellus tellus
lacinia purus, et dictum nunc justo sit amet elit.

Nullam eu ante vel est convallis dignissim. Fusce suscipit, wisi nec facilisis
facilisis, est dui fermentum leo, quis tempor ligula erat quis odio. Nunc porta
vulputate tellus. Nunc rutrum turpis sed pede. Sed bibendum. Aliquam posuere.
Nunc aliquet, augue nec adipiscing interdum, lacus tellus malesuada massa, quis
varius mi purus non odio. Pellentesque condimentum, magna ut suscipit hendrerit,
ipsum augue ornare nulla, non luctus diam neque sit amet urna. Curabitur
vulputate vestibulum lorem. Fusce sagittis, libero non molestie mollis, magna
orci ultrices dolor, at vulputate neque nulla lacinia eros. Sed id ligula quis
est convallis tempor. Curabitur lacinia pulvinar nibh. Nam a sapien.

>>>>

In this section we present a compositional representation to express the random
generation of values following the internal structure of their data types along
with the structure present on patterns matchings and abstract interfaces.

The key idea of this work is to represent different sources of structured value
information of a data type in a homogeneous way that we call a ``higher-level
representation'' (|HRep| from now on).%
\footnote{The notion of higher-level comes from that the generation process of
  given target data type is entirely determined by the type of the chosen
  representation, instead of by a concrete generator defined at the term level.}
%
For this purpose, we use a series of automatically derived data types, each one
representing an atomic unit of information that can be randomly generated and
then reflected back to the corresponding value of the original data type.

We wil reuse the previously defined data type |Exp| and the function |foo| to
explain the different concepts involved all across this section.

\subsection{\textbf{Representing Data Constructors}}

We begin by introducing the simplest representation that we can extract from our
codebase: the representation of single data constructor.
%
Each original constructor can be represented by a automatically derived data
type consisting of a single constructor with the same fields, except for the
recursive ones that are abstracted away.
%
In this light, we represent each constructor of the data type |Exp| as follows:

\begin{code}
data HRep_Val  r = Mk_Val Int
data HRep_Add  r = Mk_Add r r
data HRep_Mul  r = Mk_Mul r r
\end{code}

Note that the previous definition are type parametric over the type parameter
|r|.
%
This way we can replace |r| with any concrete data type, obtaining different
possible values on each case.
%
For instance, the value (|Mk_Add 10 20|) has type |HRep_Add Int|, while the
value (|Mk_Mul True False|) has type |HRep_Mul Bool|.
%
In practice, this parametricity let us instatiate |a| with the type of the
chosen representation (which might change according to the user demands),
without having to modify anything in the underlying machinery.

For this purpose, we employ Haskell's type classes mechanism \tocite{}.
%
We define an evaluation type class $(\Downarrow)$ which specifies how to perform
a single transformation |step| from an |HRep| back into a concrete value of the
target:
%
\begin{align*}
  \llbracket \_ \rrbracket_{target}\ :\ rep_{target}\ \rightarrow\ target
\end{align*}

% \begin{code}
% class (rep down target) where
%   step :: rep target -> target
% \end{code}

Then, we can define the following overloaded instances of the |step| operation
for the canonical |HRep| of data constructors simply by mapping each lifted
constructor back into its corresponding one.
%
\begin{alignat*}{4}
  &\evalrep{|Mk_Val|\ n\ \ \ &&}{|Exp|}
    &&= |Val|\ n \\
  &\evalrep{|Mk_Add|\ x\ y   &&}{|Exp|}
    &&= |Add|\ \evalrep{x}{|Exp|}\ \evalrep{y}{|Exp|} \\
  &\evalrep{|Mk_Mul|\ x\ y   &&}{|Exp|}
    &&= |Mul|\ \evalrep{x}{|Exp|}\ \evalrep{y}{|Exp|}
\end{alignat*}

% \begin{code}
% instance (HRep_Val down Exp) where
%   step (Mk_Val n) = Val n

% instance (HRep_Add down Exp) where
%   step (Mk_Add x y) = Add x y

% instance (HRep_Mul down Exp) where
%   step (Mk_Mul x y) = Mul x y
% \end{code}

\begin{code}
  gen_Val  gen_r  = Mk_Val  <$> arbitrary
  gen_Add  gen_r  = Mk_Add  <$> smaller gen_r <*> smaller gen_r
  gen_Mul  gen_r  = Mk_Mul  <$> smaller gen_r <*> smaller gen_r
\end{code} %$

\subsection{\textbf{Type Level Combinators}}

In order to compose the |HRep|s for the single data constructors already
presented, we will need to define a series of type level combinators.

In first place, we define a type combinator ($\otimes$) to tag a |HRep| with an
explicit generation frequency:

\begin{code}
data (f otimes n) a = Freq (f a)
\end{code}

This combinator is evaluated back to our target data type simply by piping the
result from the inner representation.
%
It does not change the evaluation semantics, as it is only considered at
generation time:
%
\begin{align*}
  \evalrep{Freq\ x : f \otimes n}{t} = \evalrep{x : f}{t}
\end{align*}

\begin{code}
gen_f_times_n gen_r = Term <$> gen_f gen_r
\end{code}

% \begin{code}
% instance (f otimes n down Exp) where
%   step (Freq f) = step f
% \end{code}

% \begin{code}
% instance Arbitrary1 (f otimes n) where
%   liftArbitrary gen_HRep = Tag <$> gen_HRep
% \end{code} %$

When generating random values, we will only allow certain terminal
representations to be generated at the moment that the generation size gets
exhausted.
%
For this, we follow the same approach as before to define a combinator (|term|)
to tag terminal representations.

\begin{code}
data (f_term) a = Term (f a)
\end{code}

\begin{align*}
  \evalrep{Term\ x : |f_term|}{t} = \evalrep{x : f}{t}
\end{align*}

\begin{code}
gen_f_term gen_r = Term <$> gen_f gen_r
\end{code}

With these individual representations, we can define a type combinator
$(\oplus)$ to compose two |HRep|s into a single one:

\begin{code}
data (f oplus g) a = L (f a) | R (g a)
\end{code}

Then, a composite representation can be transformed back into the concrete
target type by pattern matching on the data type variant and applying the |step|
tranformation to the inner |HRep|:
%
\begin{alignat*}{3}
  &\evalrep{L\ x &&: f \oplus g}{t} = \evalrep{x : f &&}{t} \\
  &\evalrep{R\ x &&: f \oplus g}{t} = \evalrep{y : g &&}{t}
\end{alignat*}

\begin{code}
  gen_f_plus_g gen_r  = sized (\size ->
    if size == 0
    then frequency
      [ (freq0  at_f,  L  <$> gen_f  gen_r  )
      , (freq0  at_g,  R  <$> gen_g  gen_r  ) ]
    else frequency
      [ (freq   at_f,  L  <$> gen_f  gen_r  )
      , (freq   at_g,  R  <$> gen_g  gen_r  ) ])
\end{code} %$

% \begin{code}
% instance (f oplus g down Exp) where
%   step (L f) = step f
%   step (R g) = step g
% \end{code}

% \begin{code}
% instance Arbitrary1 (f oplus g) where
%   liftArbitrary gen_HRep
%     = frequency
%       [ (freq at_f, L <$> gen_HRep)
%       , (freq at_g, R <$> gen_HRep) ]
% \end{code} %$

With the introduced combinators, we can easily create a type synonym |HRep_Exp|
to refer to the canonical |HRep| of our original data type |Exp|, tagging for
instance the representation of the constuctor |Add| to be generated in with
double the frequency of rest:

\begin{code}
type HRep_Exp  =       HRep_Val_term
               oplus'  HRep_Add  otimes 2
               oplus'  HRep_Mul
\end{code}

\begin{code}
  data Fix f = Fix (f (Fix f))
\end{code}

\begin{align*}
  eval : Fix\ rep \rightarrow target
\end{align*}

\subsection{\textbf{Representing Pattern Matchings}}

\begin{code}
data HRep_foo_1  r = Mk_foo_1 r r
data HRep_foo_2  r = Mk_foo_2 Int r
\end{code}

\begin{align*}
  &\evalrep{|Mk_foo_1|\ x\ y}{Exp} = \\
  &\ \ |Add (Add eval_x_Exp (Val 50)) (Add (Val 25) eval_y_Exp)|\\
  &\evalrep{|Mk_foo_2|\ x\ y}{Exp} = \\
  &\ \ |Mul (Val 50) (Mul (Val x) eval_y_Exp)|
\end{align*}

% \begin{code}
% instance (HRep_foo_1 down Exp) where
%   step (Mk_foo_1 x y)
%     = Add (Add x (Val 50)) (Add (Val 25) y)

% instance (HRep_foo_2 down Exp) where
%   step (Mk_foo_2 x y)
%     = Mul (Val 50) (Mul (Val x) y)
% \end{code}

% \begin{code}
% instance Arbitrary1 HRep_foo_1 where
%   liftArbitrary gen_HRep
%     = Mk_foo_1 <$> gen_HRep <*> gen_HRep

% instance Arbitrary1 HRep_foo_2 where
%   liftArbitrary gen_HRep
%     = Mk_foo_2 <$> arbitrary <*> gen_HRep
% \end{code} %$

\begin{code}
type HRep_foo  = HRep_foo_1 oplus''  HRep_foo_2
\end{code}

\subsection{\textbf{Representing Abstract Interfaces}}

\begin{code}
module M where

ten :: Exp
ten = Val 10

square :: Exp -> Exp
square x = Mul x x

minus :: Exp -> Exp -> Exp
minus x y = Add x (Mul y (Val (-1)))
\end{code}

\begin{code}
data HRep_ten       r = Mk_ten
data HRep_square    r = Mk_square   r
data HRep_minus     r = Mk_minus    r r
\end{code}

\begin{alignat*}{3}
  &\evalrep{|Mk_ten|         &&}{Exp} = |ten| \\
  &\evalrep{|Mk_square|\ x   &&}{Exp} = |square eval_x_Exp| \\
  &\evalrep{|Mk_minus|\ x\ y &&}{Exp} = |minus  eval_x_Exp eval_y_Exp|
\end{alignat*}

% \begin{code}
% instance (HRep_ten down Exp) where
%   step Mk_ten = ten

% instance (HRep_square down Exp) where
%   step (Mk_square x) = square x

% instance (HRep_minus down Exp) where
%   step (Mk_minus x y) = minus x y
% \end{code}

% \begin{code}
% instance Arbitrary1 HRep_ten where
%   liftArbitrary gen_HRep = pure Mk_ten

% instance Arbitrary1 HRep_square where
%   liftArbitrary gen_HRep = Mk_square <$> gen_HRep

% instance Arbitrary1 HRep_minus where
%   liftArbitrary gen_HRep
%     = Mk_minus <$> gen_HRep <*> gen_HRep
% \end{code} %$

\begin{code}
type HRep_M  = HRep_ten oplus''  HRep_square oplus''  HRep_minus
\end{code}

\begin{code}
type Spec  =       HRep_Exp  otimes 4
           oplus'  HRep_foo  otimes 2
           oplus'  HRep_M
\end{code}

This previous definition can be interpreted graphically as it is shown in the
Figure \ref{fig:hrep}.
%
Curly arrows represent the structural information extracted using
meta-programming.

\begin{figure}[t]
  \centering
  \input{tikz/hrep.tex}
  \caption{Higher level representation of the data type |Exp|, defined using
    structural information from the function |foo| and the abstract interface of
    the module |M|.}
  \label{fig:hrep}
\end{figure}

% \begin{code}
% data (Term f) a = TagTerm (f a)

% instance (Term f down Exp) where
%   step (TagTerm f) = step f
% \end{code}

### 2019-01-17 paper/sections/randomtesting.tex
WIP

In the previous definition, we declare |Exp| as the sum of three possible
classes of values: |Val| represents literal values, whereas |Add| and |Mul|
represent the addition and multiplication of two integer expressions,
respectively.
%
In Haskell, |Val|, |Add| and |Mul| are called \emph{data constructors} and are
used to distiguish which variant of the data type we are reffering to.
%
Each data constructor is then defined as a product of zero or more types known
as \emph{fields}.
%
In particular, |Val| contains a field of type |Int|, while |Add| and |Mul|
contain two fields of type |Exp| representing the operands of each operation.
%
Note that |Exp| is used as a field of at least one of its data constructors
(case |Add| and |Mul|), making it a recursively defined ADT.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this reprentation, the expression ``2 + (5 * 6)'' can be encoded simply by
using |Exp| data constructors as |(Add (Val 2) (Mul (Val 5) (Val 6)))|.
%
Furthermore, an evaluation function from |Exp|s to integer values can be defined
very idiomatically:

@@

\begin{code}
genExp :: Gen Exp
genExp = sized gen
  where
    gen 0 = Val <$> arbitrary
    gen n = oneof
      [  Val  <$> arbitrary
      ,  Add  <$> gen pred   <*> gen pred
      ,  Mul  <$> gen pred   <*> gen pred ]
\end{code} %$

This random generator is defined using \quickcheck's combinator |sized :: (Int
-> Gen a) -> Gen a| to paremetrize the generation process up to an external
integer number known as the \emph{generation size}.
%
This parameter is used to limit the maximum amount of recursive call that this
random generator can perform.
%
When called with a positive generation size (case |gen n|), the generator is free
to randomly pick with uniform probability among any data constructor of |Exp|.
%
In the case it picks to generate a |Val| data constructor, it also generates a
random |Int| value using the standard overloaded generator |arbitrary|
(\quickcheck provides standard random generators for most base types like |Int|,
|Bool|, etc.).
%
On the other hand, when it randomly picks to generate either an |Add| or a |Mul|
data constructor, it also generates independently two random subexpressions
corresponding to the data constructor fields by calling itself recursively (|gen
pred|), ensuring to decrease the generation size by a unit on each recursive
invocation.

@@

\quickcheck's default implementation for random |Int|s pick a number in the
interval $[-n, n]$ with uniform distribution.
%
Hence, in principle we need to recognize what is a suitable generation size,
which should be big enough for our generator be able to generate the numbers we
pattern match against to, otherwise we will not be able to generate any value to
match against |foo| clauses, leaving fragments of code completely untested.
%
Suppose then that we pick the a generation size $50$, i.e. the minimum size that
is big enough to produce an |Int| number equal to $50$.
%
Under this consideration, the probability of generating a value matching the
first clause of |foo| (and hence triggering the first error) results as follows:

\begin{align*}
  P(match(foo\#1))
  \!&&=\; &P(Add) \\
  \!&&*\; &P(Add) * 1 * (P(Val) * (1/100)) \\
  \!&&*\; &P(Add) * (P(Val) * (1/100)) * 1
\end{align*}

If we use \megadeth to automatically derive a random generator for |Exp|, we
obtain a uniform generation probability distribution over constructors, i.e.,
$P(|Val|) = P(|Add|) = P(|Mul|) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of |foo| only once.
%
This situation can be somewhat improved if we decide to use \dragen to obtain a
random generator.
%
Using this tool, we can optimize the generation probabilities in order to
benefit the generation of some data constructors over the rest.
%
Considering that the first pattern matching of |foo| involves the data
constructor |Add| as the only recursive one, we can set an target generation
proportion of |Add| data constructors of, for instance, $20:1$ with respect to
the rest of the generated data constructors.
%
By doing so, the obtained distribution of values results such that
$P(match(foo\#1)) \approx 1/300000$.
%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.

@@

concat :: Html -> Html -> Html
concat (Html x) (Html y) = Html (x ++ y)
\end{code}

@@

% Consider the following Haskell module |M| which defines combinators for the type
% |Exp|:

% \begin{code}
% module M where

% ten :: Exp
% ten = Val 10

% square :: Exp -> Exp
% square x = Mul x x

% minus :: Exp -> Exp -> Exp
% minus x y = Add x (Mul y (Val (-1)))
% \end{code}

>>>>

In the previous definition, we declare |Exp| as the sum of three possible
classes of values: |Val| represents literal values, whereas |Add| and |Mul|
represent the addition and multiplication of two integer expressions,
respectively.
%
In Haskell, |Val|, |Add| and |Mul| are called \emph{data constructors} (or
constructors for short) and are used to distiguish which variant of the data
type we are reffering to.
%
Each data constructor is then defined as a product of zero or more types known
as \emph{fields}.
%
In particular, |Val| contains a field of type |Int|, while |Add| and |Mul|
contain two fields of type |Exp| representing the operands of each operation.
%
Note that |Exp| is used as a field of at least one of its data constructors
(case |Add| and |Mul|), making it a recursively defined ADT.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this reprentation, the expression ``2 + (5 * 6)'' can be encoded simply by
using |Exp| data constructors as |(Add (Val 2) (Mul (Val 5) (Val 6)))|.
%
Furthermore, an evaluation function from |Exp|s to integer values can be defined
very idiomatically:

@@

\begin{code}
gen_Exp = sized (\size ->
   if size == 0
   then Val <$> arbitrary
   else frequency
     [  (1, Val  <$> arbitrary)
     ,  (2, Add  <$> smaller gen_Exp <*> smaller gen_Exp)
     ,  (1, Mul  <$> smaller gen_Exp <*> smaller gen_Exp) ])
\end{code} %$

This random generator is defined using \quickcheck's combinator |sized| to
paremetrize the generation process up to an external integer number known as the
\emph{generation size}.
%
This parameter is used to limit the maximum amount of recursive call that this
random generator can perform.
%
When called with a positive generation size (case |gen n|), the generator will
pick among any data constructor of |Exp| with a explicitly given frequency---two
times more |Add|s than |Val| or |Mul|.
%
In the case it picks to generate a |Val| data constructor, it also generates a
random |Int| value using the standard overloaded generator |arbitrary|
(\quickcheck provides standard random generators for most base types like |Int|,
|Bool|, etc.).
%
On the other hand, when it randomly picks to generate either an |Add| or a |Mul|
data constructor, it also generates independently two random subexpressions
corresponding to the data constructor fields by calling itself recursively
(|resize pred gen_Exp|), decreasing the generation size on each recursive
invocation.

@@

\quickcheck's default implementation for random |Int|s pick a number in the
interval $[-n, n]$ with uniform distribution.
%
Hence, in principle we need to recognize what is a suitable generation size,
which should be big enough for our generator be able to generate the numbers we
pattern match against to, otherwise we will not be able to generate any value to
match against |foo| clauses, leaving fragments of code completely untested.
%
Suppose then that we pick the a generation size $50$, i.e. the minimum size that
is big enough to produce an |Int| number equal to $50$.
%
Under this consideration, the probability of generating a value matching the
first clause of |foo| (and hence triggering the first error) results as follows:
%
\begin{align*}
  &P(match(foo\#1))\\
  &\quad = P(Add)       * P(Add) * 1 * (P(Val) * (1/100)) \\
  &\phantom{xxxxxxxxxx} * P(Add) * (P(Val) * (1/100)) * 1
\end{align*}

If we use \megadeth to automatically derive a random generator for |Exp|, we
obtain a uniform generation probability distribution over constructors, i.e.,
$P(|Val|) = P(|Add|) = P(|Mul|) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of |foo| only once.
%
This situation can be somewhat improved if we use \dragen to obtain a random
generator.
%
Using this tool, we can optimize the generation probabilities in order to
benefit the generation of some data constructors over the rest.
%
Considering that the first pattern matching of |foo| involves the data
constructor |Add| as the only recursive one, we can set an target generation
proportion of |Add| data constructors of, for instance, $20:1$ with respect to
the rest of the generated data constructors.
%
By doing so, the obtained distribution of values results such that
$P(match(foo\#1)) \approx 1/300000$.
%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.

@@

(<+>) :: Html -> Html -> Html
(Html x) <+> (Html y) = Html (x ++ y)
\end{code}

@@
